---
title: "Multilevel Modeling"
subtitle: "Modeling Achievement with High School and Beyond"
author: "William Murrah"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, comment=NULL)
library(lme4)
library(lmerTest)
library(performance)
library(texreg)
library(psych)
library(ggplot2)
library(car)
```


This notebook documents the use of a multilevel regression framework to model mathematics achievement using the High School and Beyond data. See the documentation for this data in the following directory in this project:

`MultilevelModelingWorkshop/resources/ho_hsb.pdf`

This project and all the necessary materials can be downloaded from the following github page:

https://github.com/auqmer/MultilevelModelingWorkshop

If you don't know how to use github, just click on the green Code tab then click "Download ZIP".

The data set can be found in `data/hsb.csv`, which contain data on 7,185 students nested within 160 schools.

### Student level variables:  
*minority* - an indicator for student ethnicity (1 = minority, 0 = other)  
*female* - an indicator for student gender (1 = female, 0 = male)  
*ses* - (a standardized scale constructed from variables measuring parental education, occupation, and income)   
*mathach* - a measure of mathematics achievement  

### School level variables:
*size* - (school enrollment)  
*sector* - (1 = Catholic, 0 = public)  
*pracad* - (proportion of students in the academic track)  
*disclim* - (a scale measuring disciplinary climate)  
*himnty* - (1 = more than 40% minority enrollment, 0 = less than 40%)   
*meanses* - (mean of the SES values for the students in this school who are included in the level-1 file)  

## Exploring HSB Data


### Load and clean the data
```{r}
# Run the R script to load and clean HSB data:
source("code/cleanHSB.R")

# Create a data set aggregated at the school level
hsb_school        <- aggregate(mathach ~ school, data = hsb, FUN = mean)
hsb_school$ses    <- aggregate(ses     ~ school, data = hsb, FUN = mean)$ses
hsb_school$sector <- aggregate(sector  ~ school, data = hsb, FUN = unique)$sector
hsb_school$size   <- aggregate(size    ~ school, data = hsb, FUN = mean)$size
```

### Visualize key student level variables

```{r}
hist(hsb$mathach, 
     breaks = "fd",
     main = "Distribution of Math Achievement Scores",
     xlab = "Math Achievement")
```
```{r}
hist(hsb$ses, 
     breaks = "fd",
     main = "Distribution of SES Scores",
     xlab = "Socioeconomic Status")
```
```{r}
plot(mathach ~ ses, 
     data = hsb,
     main = "Visualizing Relation between Math Achievement and SES",
     ylab = "Math Achievement",
     xlab = "Socioeconomic Status")
```

### Do Mean Math and SES vary by School?

```{r}

plot(mathach ~ ses, 
     data = hsb_school,
     main = "Comparing Mean Math and SES scores for each school",
     ylab = "Mean Math Scores by School",
     xlab = "Mean SES Score by School")
```

```{r, warning=FALSE}
describe(hsb, fast = TRUE)
```

```{r}

ggplot(hsb_school, aes(x = ses, y = mathach, color = sector, size = size)) +
  geom_point(alpha = .5) + labs(y = "Mean Math Achievement", 
                      x = "Mean Socioeconomic status",
                      title = "Comparing Mean Math, SES, Sector, and School Size")
```



## Models

### Empty Model

```{r}
empty.model <- lmer(mathach ~ 1 + (1|school), data = hsb)
summary(empty.model)
```

```{r}
8.614/(8.614 + 39.148)

# in the performance package:
icc(empty.model)
```

```{r}
level1.model <- lmer(mathach ~ ses + female + (1 | school), data = hsb)
summary(level1.model)
```

```{r}
level1.2.model <- lmer(mathach ~ ses + female +  
                         sector + size + disclim + (1 |school),
                       data = hsb)
summary(level1.2.model)
```

```{r}
level1.2.mses.model <- lmer(mathach ~ ses + female +  
                         sector + size + disclim + meanses + (1 |school),
                       data = hsb)
summary(level1.2.mses.model)
```

```{r}
level1.2.mses.re_ses.model <- lmer(mathach ~ ses + female +  
                         sector + size + disclim + meanses + (ses |school),
                       data = hsb)
summary(level1.2.mses.re_ses.model)
```

```{r, results='asis'}
htmlreg(list(empty.model, level1.model, level1.2.model, level1.2.mses.model,
             level1.2.mses.re_ses.model))
```

```{r}
anova(level1.2.mses.model, level1.2.mses.re_ses.model)
```


```{r}
hist(resid(empty.model), breaks = "fd")
```

```{r}
hist(resid(level1.2.mses.re_ses.model), breaks = "fd")
```

```{r}
plot(level1.2.mses.re_ses.model)
```


```{r}
plot(ranef(level1.2.mses.re_ses.model))
```


